Accurate inference of microbial differential abundance from taxonomically-biased microbiome measurements

### 2021-10-19 abundance-measurement.Rmd
Make minor edits
Edits include

- use 'widehat' instead of 'hat' over words in equations
- use 'measurement' and 'measured' instead of 'estimate' and 'estimated'
  for individual samples and change between pairs of samples, reserving
  'estimate' for regression over multiple samples
- edits based on Hypothesis comments on version
  6932587bd1ea9c56638c0ebe3ff7885b79b84e8c

Our primary tool for understanding the impact of taxonomic bias MGS measurement is the theoretical model of MGS measurement developed and empirically validated by @mclaren2019cons.
This model is the simplest model of MGS measurement that includes multiplicative taxonomic bias while respecting the *compositional* nature of sequencing measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr).
In this model, we suppose there are a set of *atomic taxa*, that differ in how efficiently they are measured but within which all cells behave similarly.
For concreteness, we equate atomic taxa with species. 
We ignore the possibility of contamination or false-positive taxonomic assignments---reads assigned to a given species and sample really do come from that species and sample.

Our model stipulates that the assigned read count of species $i$ in sample $a$ equal equals its cell density multiplied by a species-specific, sample-independent factor and a species-independent, sample-specific factor,
\begin{align}
  (\#eq:measurement-model)
  \text{reads}_{i}(a)
  = \text{density}_{i}(a) \quad \cdot
    \underbrace{\text{efficiency}_{i}}_{\substack{\text{species specific,} \\  \text{sample independent}}}
    \cdot \quad
    \underbrace{\text{sequencing effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
\end{align}
The species-specific factor is the *relative measurement efficiency* (or *efficiency* for short) of the species, which represents how much more easily that species is measured (converted from cells to assigned reads) relative to an abitrarily chosen reference species (@mclaren2019cons).
The sample-specific factor, which we call the *sequencing effort* of the sample, reflects the fact that the number of reads per unit cell density varies among samples even in the absence of taxonomic bias due to variation in total density, library normalization, and total sequencing-run output.
Equation \@ref(eq:measurement-model) implies that the total reads from all species in the sample equal
\begin{align}
  (\#eq:total-reads)
  \text{total reads}(a)
    = \text{total density}(a) \cdot \text{mean efficiency}(a) \cdot \text{sequencing effort}(a),
\end{align}
where 
\begin{align}
  (\#eq:mean-efficiency)
  \text{mean efficiency}(a) 
  \equiv \frac{\sum_{j}\text{density}_j(a)\cdot \text{efficiency}_j}{\text{total density}(a)}
\end{align}
is the average or mean efficiency of cells in the sample.

@@

The term *relative abundance* carries multiple meanings in microbiome science.
Microbiome researchers often use the relative abundance of a species in a sample to refer to its *proportion*, or number of cells from that species divided by the total number of cells from the taxonomic domain of interest (e.g. Prokaryotes).
A second view, drawn from the statistical framework of Compositional Data Analysis, equates the relative abundances of species within a sample with the *ratios* of cell densities among two or more species (e.g. a ratio of 5:4:2 among three species).
By this view, it is meaningless to refer to the relative abundance of an individual species.
Although other interpretations exist^[e.g. the rank order within a sample, or value associated with a species after a centered-log-ratio transform], we focus on the proportion and ratio views as they underly most DA analysis methods (and microbiome analysis in general), for both relative and absolute abundances, and the differences between them help to clarify the issue over when and why the effects of bias cancel in an analysis.

The standard estimate of the proportion of species $i$ in sample $a$ given by the ratio of its read count to the total,
\begin{align*}
  \hat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{total reads(a)}}.
\end{align*}
The effect of taxonomic bias on this estimate is to create a multipliative error equal to the ratio of the species' efficiency to the mean efficiency in the sample,
\begin{align}
  (\#eq:prop-error)
  \hat{\text{prop}}_{i}(a)
  &= \text{prop}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)}.
\end{align}
Intuitively, Equation \@ref(eq:prop-error) says that a species is measured as having a larger-than-actual proportion in direct relation to how much more more efficiently its cells are measured than the average cell in the sample (Figure \@ref(fig:error-proportions)).
Consequently, the same species can be over or under measured in terms of its proportion.
For instance, in samples from two hypothetical communities in Figure \@ref(fig:error-proportions), Species 3 has an efficiency of 6 and is under-estimated in Sample 1 (which has a mean efficiency of 8.33) but over-estimated in Sample 2 (which has a mean efficiency of 3.15).

The estimated ratio between species $i$ and $j$, given by the ratio of read counts, 
\begin{align*}
  \hat{\text{ratio}}_{i/j}(a) = \frac{\text{reads}_i(a)}{\text{reads}_j(a)},
\end{align*}
has a multiplicative error equal to the ratio in the species' efficiencies,
\begin{align}
  (\#eq:ratio-error)
  \hat{\text{ratio}}_{i/j}(a)
%  \frac{\text{reads}_{i}(a)}{\text{reads}_{j}(a)}
  &= \frac{\text{density}_{i}(a)}{\text{density}_{j}(a)} \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}.
\end{align}
Equation \@ref(eq:ratio-error) says that the ratio between two species is over-estimated in direct relation to how much more efficiently cells of the first species is measured relative to the second.
In contrast to estimates of proportions, estimates of species ratios have a consistent multiplicative error across samples.
For instance, in the example of Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.

@@

```{r error-proportions, out.width = '90%', fig.cap = '(ref:cap-error-proportions)', cache = TRUE}
svg_path <- here("figures/illustrations/error-proportions.svg")
if (knitr::is_html_output()) {
  knitr::include_graphics(svg_path)
} else if (knitr::is_latex_output()) {
  cowplot::ggdraw() + cowplot::draw_image(svg_path)
}
rm(svg_path)
```

(ref:cap-error-proportions) **Taxonomic bias creates sample-dependent multiplicative errors in species proportions, which can lead to inaccurate fold changes between samples.** Top row: Error in proportions measured by MGS in two microbiome samples that contain different relative abundances of three species. Bottom row: Error in the estimated fold-change in the third species that is derived from these measurements.

@@

Various experimental and computational methods have or can be used to supplement MGS measurements so as to convert the relative-abundance information in the read counts into estimates of (absolute) cell densities;
however, these methods can largely be partitioned conceptually into those that use the estimated proportions or the estimated ratios provided by MGS.
The error in density estimates due to taxonomic bias therefore follows that of the proportion and ratio estimates described above.

Proportion-based density estimation is based on the idea that if we know the (true) proportion of a species and the (true) total density of all species, we can multiply these numbers obtain the species' density.
This method therefore involves multiplying the proportion of a species estimated from MGS with an estimate of total microbial density obtain through other means (e.g., through direct cell counting or qPCR of a universal marker gene) to estimate 
its density,
\begin{align}
  (\#eq:density-prop-est)
  \hat{\text{density}}_{i}(a) 
  = \hat{\text{prop}}_{i}(a) \cdot \hat{\text{total density}}(a).
\end{align}
The error in this proportion-based density estimate is
\begin{align}
  (\#eq:density-prop-error)
  \hat{\text{density}}_{i}(a) 
  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)} 
  \cdot \substack{\text{fold error in} \\ \hat{\text{total density}}(a)}.
\end{align}
The multiplicative error consists of two factors: the error in the estimated proportion (due to taxonomic bias in the MGS measurement), and any error in the estimated total density.
In the case of perfect knowledge of the total density, the fold error in proportion-based density estimates varies across samples inversely with the mean efficiency.

Ratio-based density estimation follows from the fact that, in the absence of taxonomic bias, the conversion rate from reads to density is just the sequencing effort of the sample and is thus the same for all species (Equation \@ref(eq:measurement-model)).
Knowing the density of at least one _reference_ species therefore enables us to convert from reads to density for all species.
One approach to obtaining such reference species is to add one or more extraneous species in known (typically constant) abundance (a "spike-in") to each sample prior to sequencing (REFs).
A second approach is to determine a set of naturally occurring species that are thought to have a constant, though typically unknown, abundance across samples; in this case we can only estimate species' densities relative to these "housekeeping species" (REFs).
A third (seemingly untried) possibility is to measure the density of one or more naturally occurring species using a method such as targeted CFU counting, fluorescence flow cytometry or ddPCR directly on cells, or ddPCR or qPCR on extracted DNA.
Given the estimate of the density of reference taxon $r$, we estimate the density of taxon $i$ by multipling the ratio of $i$ to $r$ in the reads by the density of $r$,
\begin{align}
  (\#eq:density-ratio-est)
  \hat{\text{density}}_{i}(a) = \frac{\text{reads}_{i}(a)}{\text{reads}_{r}(a)} \cdot \hat{\text{density}}_{r}(a)
\end{align}
(Appendix ??? describes how this equation can be extended to multiple reference species.)
For a constant reference species with unknown density, we can set $\hat{\text{density}}_{r}(a)$ equal to 1 to estimate density in units of the density of the reference species.
The error in this ratio-based density estimate is
\begin{align}
  (\#eq:density-ratio-error)
  \hat{\text{density}}_{i}(a) 
  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}} 
  \cdot \substack{\text{fold error in} \\ \hat{\text{density}}_{r}(a)}.
\end{align}
Again, the multiplicative error consists of two factors: the error in the estimated ratio (due to taxonomic bias in the MGS measurement), and error in the estimated density of the reference taxa.
So long as the fold error in the reference density is constant across samples, so will be the error in the estimated density of the focal species $i$.

- Note: The current distinction between proportion- and ratio-based density estimation is incomplete, as the example of spike-in normalization shows: Whether the density of species $i$ is estimated by Equation \@ref(eq:density-ratio-est), or by first estimating total density from the ratio of non-spike-in to spike-in reads and then using Equation \@ref(eq:density-prop-est), the results are mathematically identical. Future revisions should clarify this point.

>>>>

## Model of MGS measurement

Our primary tool for understanding the impact of taxonomic bias MGS measurement is the theoretical model of MGS measurement developed and empirically validated by @mclaren2019cons.
This model is the simplest model of MGS measurement that includes multiplicative taxonomic bias while respecting the *compositional* nature of sequencing measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr).
We consider a set of microbial species that are measured by a specific MGS protocol. The protocol assigns each sequencing read to at most one sample andsample and  species, and we ignore the possibility of contamination or false-positive taxonomic assignments---reads assigned to a given species and sample really do come from that species and sample.
Our model stipulates that the assigned read count of species $i$ in sample $a$ equals its cell density multiplied by a species-specific, sample-independent factor and a species-independent, sample-specific factor,
\begin{align}
  (\#eq:measurement-model)
  \text{reads}_{i}(a)
  = \text{density}_{i}(a) \quad \cdot
    \underbrace{\text{efficiency}_{i}}_{\substack{\text{species specific,} \\  \text{sample independent}}}
    \cdot \quad
    \underbrace{\text{sequencing effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
\end{align}
The species-specific factor is the *relative measurement efficiency* (or *efficiency* for short) of the species, which represents how much more easily that species is measured (converted from cells to assigned reads) relative to an abitrarily chosen reference species (@mclaren2019cons).
The sample-specific factor, which we call the *sequencing effort* of the sample, reflects the fact that the number of reads per unit cell density varies among samples even in the absence of taxonomic bias due to variation in total density, library normalization, and total sequencing-run output.
Equation \@ref(eq:measurement-model) implies that the total reads from all species in the sample equal
\begin{align}
  (\#eq:total-reads)
  \text{total reads}(a)
    = \text{total density}(a) \cdot \text{mean efficiency}(a) \cdot \text{sequencing effort}(a),
\end{align}
where 
\begin{align}
  (\#eq:mean-efficiency)
  \text{mean efficiency}(a) 
  \equiv \frac{\sum_{j}\text{density}_j(a)\cdot \text{efficiency}_j}{\text{total density}(a)}
\end{align}
is the average or mean efficiency of cells in the sample.

@@

The term *relative abundance* carries multiple meanings in microbiome science.
Microbiome researchers often use the relative abundance of a species in a sample to refer to its *proportion*, or number of cells from that species divided by the total number of cells from the taxonomic domain of interest (e.g. Prokaryotes).
A second view, drawn from the statistical framework of Compositional Data Analysis, equates the relative abundances of species within a sample with the *ratios* of cell densities among two or more species (e.g. a ratio of 5:4:2 among three species).
By this view, it is meaningless to refer to the relative abundance of an individual species.
Although other interpretations exist^[e.g. the rank order within a sample, or value associated with a species after a centered-log-ratio transform], we focus on the proportion and ratio views as they underly most relative  and absolute DA methods and the differences between them clarify when and why the effects of bias cancel in a DA analysis.

The standard measurement of the proportion of species $i$ in sample $a$ is given by the ratio of its read count to the total,
\begin{align}
  (\#eq:prop-meas)
  \widehat{\text{prop}}_{i}(a) = \frac{\text{reads}_i(a)}{\text{total reads(a)}}.
\end{align}
From Equations \@ref(eq:measurement-model), \@ref(eq:total-reads), and \@ref(eq:prop-meas), it follows that
\begin{align}
  (\#eq:prop-error)
  \widehat{\text{prop}}_{i}(a)
  &= \text{prop}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)},
\end{align}
where $\text{prop}_{i}(a) = \text{density}_{i}(a) / \text{total density}(a)$ is the actual proportion.
Equation \@ref(eq:prop-error) says that taxonomic bias creates a multiplicative error equal to the ratio of the species' efficiency to the mean efficiency in the sample.
Consequently, the same species can be over- or under-measured depending on the composition of the given sample (Figure \@ref(fig:error-proportions)).
For instance, in samples from two hypothetical communities in Figure \@ref(fig:error-proportions), Species 3 has an efficiency of 6 and is under-measured in Sample 1 (which has a mean efficiency of 8.33) but over-measured in Sample 2 (which has a mean efficiency of 3.15).

The measured ratio between species $i$ and $j$ is given by the ratio of their read counts, 
\begin{align}
  (\#eq:ratio-meas)
  \widehat{\text{ratio}}_{i/j}(a) = \frac{\text{reads}_i(a)}{\text{reads}_j(a)}.
\end{align}
From Equations \@ref(eq:measurement-model) and \@ref(eq:ratio-meas), it follows that 
\begin{align}
  (\#eq:ratio-error)
  \widehat{\text{ratio}}_{i/j}(a)
%  \frac{\text{reads}_{i}(a)}{\text{reads}_{j}(a)}
  &= \frac{\text{density}_{i}(a)}{\text{density}_{j}(a)} \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}.
\end{align}
The fold error in the ratio between two species simply equals the ratio in their efficiencies and is therefore constant across samples.
For instance, in the example of Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.

@@

```{r error-proportions, out.width = '90%', fig.cap = '(ref:cap-error-proportions)', cache = TRUE}
svg_path <- here("figures/illustrations/error-proportions.svg")
if (knitr::is_html_output()) {
  knitr::include_graphics(svg_path)
} else if (knitr::is_latex_output()) {
  # cowplot::ggdraw() + cowplot::draw_image(svg_path)
  pdf_path <- file.path("/tmp", "bookdown-differential-abundance-theory",
    "error-proportions.pdf")
  args <- c(
    "--export-area-page", 
    "--export-type=pdf", 
    str_c("--export-filename=", pdf_path),
    svg_path
  )
  system2("inkscape", args = args)
  knitr::include_graphics(pdf_path)
  rm(pdf_path)
}
rm(svg_path)
```

(ref:cap-error-proportions) **Taxonomic bias creates sample-dependent multiplicative errors in species proportions, which can lead to inaccurate fold changes between samples.** Top row: Error in proportions measured by MGS in two microbiome samples that contain different relative abundances of three species. Bottom row: Error in the measured fold-change in the third species that is derived from these measurements.

@@

A range of experimental and computational methods can be used to convert the relative-abundance information in the read counts into measurements of (absolute) cell densities;
however, these methods can largely be partitioned into those that use the measured proportions or the measured ratios provided by MGS.
The error in density measurements due to taxonomic bias therefore follows that of the proportion and ratio measurements described above.

Proportion-based density measurement is based on the idea that if we know the (true) proportion of a species and the (true) total density of all species, we can multiply these numbers obtain the species' density.
This method therefore involves multiplying the proportion of a species measured from MGS with a measurement of total microbial density obtain through other means (e.g., through direct cell counting or qPCR of a universal marker gene) to measure its density,
\begin{align}
  (\#eq:density-prop-meas)
  \widehat{\text{density}}_{i}(a) 
  = \widehat{\text{prop}}_{i}(a) \cdot \widehat{\text{total density}}(a).
\end{align}
Substituting the expression \@ref(eq:prop-error) for the measured proportion and allowing for error in the total density measurement, leads to the expression
\begin{align}
  (\#eq:density-prop-error)
  \widehat{\text{density}}_{i}(a) 
  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{mean efficiency}(a)} 
  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{total density}}(a) \end{array}.
\end{align}
The multiplicative error consists of two factors: the error in the estimated proportion (due to taxonomic bias in the MGS measurement), and any error in the estimated total density.
In the case of perfect knowledge of the total density, the fold error in proportion-based density estimates varies across samples inversely with the mean efficiency.

Ratio-based density estimation follows from the fact that, in the absence of taxonomic bias, the conversion rate from reads to density is just the sequencing effort of the sample and is thus the same for all species (Equation \@ref(eq:measurement-model)).
Knowing the density of at least one _reference_ species therefore enables us to convert from reads to density for all species.
One approach to obtaining such reference species is to add one or more extraneous species in known (typically constant) abundance (a "spike-in") to each sample prior to sequencing (REFs).
A second approach is to determine a set of naturally occurring species that are thought to have a constant, though typically unknown, abundance across samples; in this case we can only estimate species' densities relative to these "housekeeping species" (REFs).
A third (seemingly untried) possibility is to measure the density of one or more naturally occurring species using a method such as targeted CFU counting, fluorescence flow cytometry or ddPCR directly on cells, or ddPCR or qPCR on extracted DNA.
Given the estimate of the density of reference taxon $r$, we estimate the density of taxon $i$ by multipling the ratio of $i$ to $r$ in the reads by the density of $r$,
\begin{align}
  (\#eq:density-ratio-meas)
  \widehat{\text{density}}_{i}(a) = \frac{\text{reads}_{i}(a)}{\text{reads}_{r}(a)} \cdot \widehat{\text{density}}_{r}(a)
\end{align}
(Appendix ??? describes how this equation can be extended to multiple reference species.)
For a constant reference species with unknown density, we can set $\widehat{\text{density}}_{r}(a)$ equal to 1 to estimate density in units of the density of the reference species.
The error in this ratio-based density estimate is
\begin{align}
  (\#eq:density-ratio-error)
  \widehat{\text{density}}_{i}(a) 
  = \text{density}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}} 
  \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{density}_r}(a) \end{array}.
\end{align}
Again, the multiplicative error consists of two factors: the error in the estimated ratio (due to taxonomic bias in the MGS measurement), and error in the estimated density of the reference taxa.
So long as the fold error in the reference density is constant across samples, so will be the error in the estimated density of the focal species $i$.

- Note: The current distinction between proportion- and ratio-based density estimation is incomplete, as the example of spike-in normalization shows: Whether the density of species $i$ is estimated by Equation \@ref(eq:density-ratio-meas), or by first estimating total density from the ratio of non-spike-in to spike-in reads and then using Equation \@ref(eq:density-prop-meas), the results are mathematically identical. Future revisions should clarify this point.
